\documentclass[twoside]{MATH77}
\begin{document}
\begcover
\begin{center}
\begin{minipage}{5.5in}
\vspace{1.4in}
\hspace{1.5in}You can only go forward\vspace{10pt}

\hspace{1.75in}by facing the obligation\vspace{10pt}

\hspace{2in}of mastering the weight\vspace{10pt}

\hspace{1in}of what the titans of the idiom have laid down.\vspace{10pt}

\hspace{2.25in}---  Wynton Marsalis

\if j\outtyp
\vspace{1in}
Second printing, with corrections.  June~12, 1992.\vspace{10pt}

Third printing, with corrections in Chapters 8.2, 8.3, and~9.3.
September~23, 1992.\vspace{10pt}

Fourth printing, with corrections in Chapters 1, 2.2, 2.4, 2.5, 9.2,
16.4, 18.1, 18.2, 19.2 and~19.3, additions to Chapter~2.15, revision
of Appendices A, B and C and redesignation to be Appendices E, A and
B, respectively, and addition of Chapters 2.18, 2.19, 11.5, 11.6,
22.2, 22.3, 22.4 and Appendices C and D.  July~27, 1993.\vspace{10pt}

Fifth printing, text converted from WordPerfect 5.1 to \TeX \ format.
Revisions and corrections throughout the document, addition of a
single precision C library, addition of Chapter~19.4, and addition of
Hermite interpolation in Chapter~12.1. January~18, 1995.\vspace{10pt}

Sixth printing, documentation converted from \LaTeX\ to \LaTeX2e, chapter
numbers changed on a few chapters, references converted to use \BibTeX,
dropped Appendix E which gave correspondence of routines here with the old
JPL\$ library, and miscellaneous corrections in the documentation.
Conversion of Fortran codes to C, the checking of results on different
machines, and the generation of the Index and Appendices A, B, and D are
now automated. New Chapters are:\\[-17pt]
\fi
\end{minipage} \end{center}
\if j\outtyp
\begin{tabbing}
\hspace{1.in}\=\hspace{.5in}\=\kill
\> \phantom{1}2.20 \> Binomial Coefficients\\
\> 12.4 \> C$^0$ and C$^1$ Surface Interpolation to Scattered Data\\
\> 12.5 \> Least Squares Fit of Bicubic Spline Surface\\
\> 13.0 \> Effective Use of the Quadrature Software\\
\> 14.2 \> Explicit Runge-Kutta Method for Ordinary~Differential~Equations\\
\> 16.3 \> Plotting Using \TeX \\
\> 19.5 \> Checking the Installed Library\\
\> 19.6 \> Testing Functions of One and Two Variables\\
\> 19.7 \> Checking and Output of Program Unit Interfaces\\[-25pt]
\end{tabbing}
\hspace{.75in}May 23, 1998
\fi
\newpage
\end{document}
